Dissertation Proposal

### 2023-11-08 proposal.tex
grammar

The current scope tutorial uses the following prompt:
\begin{quote}
  When I see a variable reference, how do I find the corresponding
  declaration, if any?
\end{quote}
This prompt has a few problems:
\begin{itemize}
  \item It is unclear to adapt it to other tutorials.
  \item Students tend to give brief responses possibly because the
  questions is asking for a lot and because students are tired when
  they reach the end of a tutorial.
\end{itemize}

So I plan to change the question format into a notice-and-wonder style
of questions, which is known to be less demanding
(\Cref{s:pedagogic-techniques}). So I plan to replace the
\textbf{scope} prompt with
\begin{quote}
  \emph{Think about} the following questions:
  \begin{itemize}
    \item How do \lstinline|defvar|s work?
    \item How do \lstinline|deffun|s work?
    \item How do variable referencess (e.g., \lstinline|x| in \lstinline|(+ x 1)|) work?
    \item How do \lstinline|defvar|s, \lstinline|deffun|s, and variable references interact with each other?
  \end{itemize}
  If you believe you know the answers perfectly, please \emph{response with}
  \begin{quote}
    I know the answers with 100\% certainty.
  \end{quote}
  Otherwise, please \emph{response with} what you are uncertain about these language constructs.
\end{quote}
I plan to use a similar prompt in later tutorials. For example, in \textbf{mut-vars},
I would want the tutor to ask
\begin{quote}
  \emph{Think about} the following questions:
  \begin{itemize}
    \item How do \lstinline|set!|s work?
    \item How do \lstinline|set!|s interact with other language constructs (namely, \lstinline|defvar|s, \lstinline|deffun|s, and variable references)?
  \end{itemize}
  If you believe you know the answers perfectly, please \emph{response with}
  \begin{quote}
    I know the answers with 100\% certainty.
  \end{quote}
  Otherwise, please \emph{response with} what you are uncertain about these language constructs.
\end{quote}
I am hoping these new prompt are

@@

My plan is to change the tutor such that it provides a
program-specific link after each interpreting task and instructs
students to click the link for more explanation.

@@

To fully answer my research questions, I should also argue that my
tutor is effective at correcting misconceptions. In this section, I
describe how I want to refine the design of the tutor to help me make
the argument, what data I want to collect, and what (statistical)
analysis do I want to perform on the data.

I can infer whether students hold a misconception by analyzing their
responses to interpreting tasks. There are three possible responses to
an interpreting task:
\begin{enumerate}
  \item They give the correct answer.
  \item They give a wrong answer that correspond to some
        misconceptions.
  \item They give a wrong answer that do not correspond to any (known)
        misconceptions.
\end{enumerate}
If a student gives a wrong answer corresponds to exactly one
misconception \emph{and} an explanation that matches the
misconception, I deduce that the student holds the misconception when
doing the task.

@@

The current tutor is not in the final shape yet. In particular, many
of the wrong answers correspond to more than one misconception; some
misconceptions are represented by only one question; the order of
interpreting tasks are not randomized. I plan to tackle the first two
differences by adding and removing programs. Adding randomization
requires me to carefully check the dependencies between current
explanations and other texts in the tutor and resolve those
dependencies if any.

@@

\emph{Case comparisons} draws analogies between examples.
\cite{alfieriLearningCaseComparisons2013a} suggests that asking
(rather than not asking) students to find similarities between cases,
and providing principles \emph{after} the comparisons (rather than
before or not at all), are associated with better learning outcomes.

Several groups of authors have argued for the benefits of visualizing
program executions
\citep{dicksonExperiencesImplementingUtilizing2022,
guoOnlinePythonTutor2013, karnalimUsePythonTutor2017,
napsExploringRoleVisualization2002}. These visualization seems to
increase engadgments and help students understand how programs works.
These work often mention \emph{notional machines}
\citep{duboulayBlackBoxGlass1981}, which refers to the underlying
mental models rather than the visual representations.

Prior research argues for the benefits of \emph{language levels}
\citep{findlerDrSchemeProgrammingEnvironment2002,crestaniExperienceReportGrowing2010}:
a langauge has better be taught as a sequence of languages, where each
later language includes more langauge constructs than the previous
one.

@@

I therefore view the above data as
purely formative: they suggest that the Tutor \emph{most likely did
  not do harm} and perhaps even \emph{may have done some
  good}. However, it would be improper to read too much into the
analysis. It is quite possible that some of the other problems would
have found issues. Rather, what I really see is value in the
misinterpreter concept. It is not only useful for analysis, it is also
valuable for \emph{problem design}: in the next iteration of the
Tutor, I will use misinterpreters actively to shape the incorrect answers and,
as necessary, update the programs as well. I therefore hope to have
much more thorough analyses in future work.

@@

Fortunately, I were able to deploy the Tutor on two other
populations: University 2 and Texbook (See \Cref{s:populations} for their details).
% \begin{itemize}

@@

By the end of Fall 2023, I would like to improve the SMoL Tutor:
\begin{enumerate}
  \item Ask students to explain when they gave a wrong answer in an
        interpreting question. With student explanations, I check whether
        choosing the wrong answer indicates holding the corresponding misconception(s).
  \item Adjust the collection of interpreting questions such that
        every misconception has at least two questions that aims at
        correcting the misconception.
  \item Partially randomize the order of questions.
\end{enumerate}

@@

During Fall 2024, I plan to collect a new round of data from Brown. I
expect these datasets and the datasets from Spring 2024 would give a
more rigorously evaluation of the effectiveness of SMoL Tutor. The
improvements to be made in Fall 2023 should resolve many of the
threats to validity. In this semester I also want to write a new paper
about the SMoL Tutor.

I plan to defense in Spring 2025.

>>>>

The current scope tutorial uses the following prompt:
\begin{quote}
  When I see a variable reference, how do I find the corresponding
  declaration, if any?
\end{quote}
This prompt has a few problems:
\begin{itemize}
  \item It is unclear to adapt it to other tutorials.
  \item Students tend to give brief responses possibly because the
  questions are asking for a lot and because students are tired when
  they reach the end of a tutorial.
\end{itemize}

So I plan to change the question format into a notice-and-wonder style
of questions, which is known to be less demanding
(\Cref{s:pedagogic-techniques}). So I plan to replace the
\textbf{scope} prompt with
\begin{quote}
  \emph{Think about} the following questions:
  \begin{itemize}
    \item How do \lstinline|defvar|s work?
    \item How do \lstinline|deffun|s work?
    \item How do variable referencess (e.g., \lstinline|x| in \lstinline|(+ x 1)|) work?
    \item How do \lstinline|defvar|s, \lstinline|deffun|s, and variable references interact with each other?
  \end{itemize}
  If you believe you know the answers perfectly, please \emph{response with}
  \begin{quote}
    I know the answers with 100\% certainty.
  \end{quote}
  Otherwise, please \emph{respond with} what you are uncertain about these language constructs.
\end{quote}
I plan to use a similar prompt in later tutorials. For example, in \textbf{mut-vars},
I would want the tutor to ask
\begin{quote}
  \emph{Think about} the following questions:
  \begin{itemize}
    \item How do \lstinline|set!|s work?
    \item How do \lstinline|set!|s interact with other language constructs (namely, \lstinline|defvar|s, \lstinline|deffun|s, and variable references)?
  \end{itemize}
  If you believe you know the answers perfectly, please \emph{respond with}
  \begin{quote}
    I know the answers with 100\% certainty.
  \end{quote}
  Otherwise, please \emph{respond with} what you are uncertain about these language constructs.
\end{quote}

@@

My plan is to change the tutor so that it provides a
program-specific link after each interpreting task and instructs
students to click the link for more explanation.

@@

To fully answer my research questions, I should also argue that my
tutor is effective at correcting misconceptions. In this section, I
describe how I want to refine the design of the tutor to help me make
the argument, what data I want to collect, and what (statistical)
analysis I want to perform on the data.

I can infer whether students hold a misconception by analyzing their
responses to interpreting tasks. There are three possible responses to
an interpreting task:
\begin{enumerate}
  \item They give the correct answer.
  \item They give a wrong answer that corresponds to some
        misconceptions.
  \item They give a wrong answer that does not correspond to any (known)
        misconceptions.
\end{enumerate}
If a student gives a wrong answer that corresponds to exactly one
misconception \emph{and} an explanation that matches the
misconception, I deduce that the student holds the misconception when
doing the task.

@@

The current tutor is not in the final shape yet. In particular, many
of the wrong answers correspond to more than one misconception; some
misconceptions are represented by only one question; the order of
interpreting tasks is not randomized. I plan to tackle the first two
differences by adding and removing programs. Adding randomization
requires me to carefully check the dependencies between current
explanations and other texts in the tutor and resolve those
dependencies, if any.

@@

\emph{Case comparisons} draw analogies between examples.
\cite{alfieriLearningCaseComparisons2013a} suggest that asking
(rather than not asking) students to find similarities between cases,
and providing principles \emph{after} the comparisons (rather than
before or not at all), are associated with better learning outcomes.

Several groups of authors have argued for the benefits of visualizing
program executions
\citep{dicksonExperiencesImplementingUtilizing2022,
guoOnlinePythonTutor2013, karnalimUsePythonTutor2017,
napsExploringRoleVisualization2002}. Visualization seems to
increase engagement and help students understand how programs work.
These work often mention \emph{notional machines}
\citep{duboulayBlackBoxGlass1981}, which refers to the underlying
mental models rather than the visual representations.

Prior research argues for the benefits of \emph{language levels}
\citep{findlerDrSchemeProgrammingEnvironment2002,crestaniExperienceReportGrowing2010}:
a language has better be taught as a sequence of languages, where each
later language includes more language constructs than the previous
one.

@@

I therefore view the above data as
purely formative: they suggest that the Tutor \emph{most likely did
  not do harm} and perhaps even \emph{may have done some
  good}. However, it would be improper to read too much into the
analysis. It is quite possible that some of the other problems would
have found issues. Rather, what I really see is value in the
misinterpreter concept. It is not only useful for analysis, it is also
valuable for \emph{problem design}: in the next iteration of the
Tutor, I will use misinterpreters actively to shape the incorrect answers and,
as necessary, update the programs as well. I therefore hope to have
much more thorough analyzes in future work.

@@

Fortunately, I was able to deploy the Tutor on two other
populations: University 2 and Texbook (See \Cref{s:populations} for their details).
% \begin{itemize}

@@

By the end of Fall 2023, I would like to improve the SMoL Tutor:
\begin{enumerate}
  \item Ask students to explain when they gave a wrong answer in an
        interpreting question. With student explanations, I check whether
        choosing the wrong answer indicates holding the corresponding misconception(s).
  \item Adjust the collection of interpreting questions such that
        every misconception has at least two questions that aim at
        correcting the misconception.
  \item Partially randomize the order of questions.
\end{enumerate}

@@

During Fall 2024, I plan to collect a new round of data from Brown. I
expect these datasets and the datasets from Spring 2024 would give a
more rigorous evaluation of the effectiveness of SMoL Tutor. The
improvements to be made in Fall 2023 should resolve many of the
threats to validity. In this semester I also want to write a new paper
about the SMoL Tutor.

I plan to defend in Spring 2025.
